hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-04 inst/manuscript/paper.Rmd
final copy edits and render 2

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

**hatchR** is designed primarily as a tool for predicting fish phenology. To maintain focus on this core function, we provide minimal built-in data validation and vizualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensize validation checks. However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`). Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function. Users can then apply these models to water temperature data (*e.g.,* from a HOBO temperature logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external visualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows–particularly for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.

@@

This framework as the foundation for phenological models in **hatchR**. The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979). While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function. This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond salmonids.

@@

To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically. These checks ensure data integroty before model application.

@@

To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd). Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3). This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.

@@

To illustrate model selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**. Speficially, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.

@@

The output from `predict_phenology()` contains mutiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:

@@

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population. Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walkthrough is available on the **hatchR** Shiny website.

@@

For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes. Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022]. These findings highlight the importance of considering how developmental rates are keyedto specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.

@@

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature data sets.

>>>>

This effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

**hatchR** is designed primarily as a tool for predicting fish phenology. To maintain focus on this core function, we provide minimal built-in data validation and visualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensive validation checks. However, we include basic data-checking and summarization functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`). Intuitive functions are provided for users to apply models—either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function. Users can then apply these models to water temperature data (*e.g.,* from a HOBO temperature logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external visualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows–particularly for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.

@@

This framework as the foundation for phenological models in **hatchR**. The package includes a predefined `model_table` containing established parameterizations, including those from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979). While `model_table` incorporated more complex models from @beacham1990, users can also fit custom models using the `fit_model()` function. This flexibility allows for the incorporation of new parameterizations as they are developed, expanding the utility of **hatchR** beyond Salmonids.

@@

To help users identify potential issues, **hatchR** includes basic data checking functions that highlight outliers or missing values both visually and programatically. These checks ensure data integrity before model application.

@@

To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd). Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3). This example highlights the flexibility of **hatchR** for accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.

@@

To illustrate model selection and phenology prediction, we will replicate a portion of the analysis from @sparks2019 using the `woody_island` dataset included with **hatchR**. Specifically, we predict both hatch and emergence timing for Sockeye Salmon at Woody Island in 1990.

@@

The output from `predict_phenology()` contains multiple elements in a list, which can be accessed using the `$` operator. Each component provides different insight into the predicted phenology:

@@

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population. Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensitive to sediment disturbance. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walk through is available on the **hatchR** Shiny website.

@@

For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes. Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022]. These findings highlight the importance of considering how developmental rates are keyed to specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.

@@

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided helpful discussion about model development and temperature data sets.
